A lottery model of density-dependent selection in evolutionary genetics

### 2017-09-22 manuscript.tex
latest

Fitness is typically represented in heavily simplified terms in evolutionary genetics, often using constant selection coefficients, to make it easier to infer or predict how type frequencies change over time. This excludes fundamental ecological factors such as dynamic population size or density-dependence from the most genetically-realistic treatments of evolution, a problem that inspired MacArthur's influential but problematic $r$/$K$ theory. Following the spirit of $r$/$K$-selection as a general-purpose theory of density-dependent selection, we develop a new model of density-dependent selection by generalizing the fixed-density classic lottery model of territorial acquisition to accommodate arbitrary population densities. We show that, with density dependence, co-existence is possible in the lottery model in a stable environment. Inspired by natural \textit{Drosophila} populations, we consider co-existence under strong, seasonally-fluctuating selection coupled to large cycles in population density, and show that co-existence (stable polymorphism) is promoted via a combination of the classic storage effect and density-regulated population growth. We also show that the only significant bias introduced by selection at the different environmental extremes of Grime's triangle is the relative importance of competitive ability at different densities, confirming an important role for phenotypic constraints in shaping ``primary strategies''.

@@

``...the concept of fitness is probably too complex to allow of a useful mathematical development. Since it enters fundamentally into many population genetics considerations, it is remarkable how little attention has been paid to it.'' --- Warren J. Ewens, Mathematical Population Genetics I, 2004, pp. 276

@@

Evolutionary models differ greatly in their treatment of fitness. In models of genetic evolution, genotypes are typically assigned constant (or occasionally frequency-dependent) selection coefficients describing the change in their relative frequencies over time. This simplified treatment of selection facilitates explicit time-dependent treatment of genotype frequencies, and can be justified over sufficiently short time intervals \citep[p. 276]{ewens_2004}. The emphasis of population genetics is to infer past selection, migration and demographic change given a sample of nucleotide sequences, or to predict how allele frequencies change over time based on their relative fitness effects together with population structure, genetic drift and linkage. The resulting picture of evolution excludes basic elements of the ecological underpinnings of selection, including density dependence, and how selection affects population size. This complicates the inference of past selection, because demographic changes can look genealogically very similar to selective frequency changes \citep{barton_1998}.

@@

It is challenging to generalize beyond particular traits or ecological scenarios to model fundamentally different forms of selection. Perhaps this is not surprising given that fitness is such a complex quantity, dependent on all of a phenotype's functional traits \citep{violle_2007} and its environment. A detailed, trait-based, predictive model of fitness would be enormously complicated and situation-specific. It is therefore easy to doubt the feasibility of a simplified, general mathematical treatment of fitness \citep[p. 276]{ewens_2004}. For example, MacArthur's famous $r$/$K$ scheme \citep{macarthur_1962,macarthur_1967} is now almost exclusively known as a framework for understanding life-history traits, and judged on its failure in that role \citep{pianka_1970,stearns_1977,boyce_1984,reznick_2002}. However, the r/K scheme's original purpose was to extend the existing population-genetic treatment of selection to account for population density \citep{macarthur_1962}. Few attempts have been made to develop it further along these lines.

However, empirical trait classification studies have suggested the existence of a few ``primary strategies'', reflecting broadly distinct responses to selection \citep{winemiller_2015}. Grime famously considered \citep{grime_1974,grime_1977,grime_1988,westoby_1998} two broad determinants of population density: stress (persistent hardship e.g. due to resource scarcity or unfavorable temperatures) and disturbance (intermittent destruction of vegetation e.g. due to trampling, herbivory, pathogens, extreme weather or fire).  The extremes of these two factors define three primary strategies denoted by C/S/R respectively (Fig. \ref{fig:axes}): competitors ``C'' excel in low stress, low disturbance environments; stress tolerators ``S'' excel in high stress, low disturbance environments; and ruderals ``R''  excel in low stress, high disturbance environments. Population persistence is not possible in high-stress, high-disturbance environments. Grime showed that measures of C, S and R across a wide range of plant species are anti-correlated, so that strong C-strategists are weak S and R strategists, and so on, creating a triangular C/S/R ternary plot \citep{grime_1974}. Similar schemes were proposed for insects \citep{southwood_1977}, fishes \citep{winemiller_1992}, and zooplankton \citep{allan_76}. More recently, modern hierarchical clustering techniques have also revealed, with greater statistical rigor, distinct trait clusters in corals analogous to Grime's primary strategies \citep{darling_2012}. These empirical findings suggest that functional traits contribute to fitness predominantly via a few key factors such as intrinsic reproductive rate or stress-tolerance.

\begin{figure}
\centering
\includegraphics[scale=1]{axes.pdf}
\caption{\label{fig:axes} Schematic of Grime's triangle and the associated direction of selection in our model. The two axes show increasing levels of environmental stress and disturbance. Population persistence is not possible if the combination of stress and disturbance is too large (below dashed line). This creates a triangle, each corner of which corresponds to an environmental extreme and associated ``primary strategy''. The high-density (HD) interpretation of stress is shown as a rotation of the triangle with respect to density. The strength of selection on $c$ depends on density $L$, but $b$ and $d$ experience the same selective force in all three environmental extremes in both versions of the triangle.} 
\end{figure}

Here we explore the interplay between some ``key factors'' of fitness in a simplified, territorial model of growth, dispersal and competition. This broadly follows the original spirit of MacArthur's $r/K$ scheme. More specifically, our aim is to begin merging some ecological realism into population genetics' time-dependent, genetically-focused view of evolution. We revisit the classic lottery model of \cite{chesson_1981}, which has two features that make it well suited for this role, but one critical flaw that we rectify here.

@@

Using our extended lottery model, we then revisit Grime's C/S/R scheme, and evaluate the extent to which selection favors fecundity, competitive ability or lower mortality under Grime's environments (section ``Primary strategies and Grime's triangle''). This represents a simple ``sanity check'' on the invasion behavior of our extended lottery model under different extremes.

We then explore some time-dependent behavior of our extended lottery model. Taking an example inspired by recent studies of rapid, seasonal evolution in \textit{Drosophila} \citep{bergland_14}, we discuss how environmental fluctuations might stabilize polymorphisms when population density is cyclical.

@@

\subsection*{Primary strategies and Grime's triangle}

Here we describe how Grime's ``disturbed'', ``stressful'' and ``ideal'' environments can be captured mathematically in our model, and evaluate the role of selection in shaping the corresponding ``primary strategies''. To proceed, we map these verbally-defined environments to quantitative parameter regimes in our model, summarized in Fig.~\ref{fig:table}.

Ideal environments are characterized by the near-absence of stress and disturbance, so that survival is easy ($d\ll 1$) and growth and reproduction can be rapid ($b\gg d$). The propagule density is determined by the magnitude of $b/d$, and hence $L\gg 1$ (Appendix C).

Disturbed environments are characterized by short bursts of high extrinsic mortality $d$ caused by physical destruction. We assume that disturbances are equally damaging to adults and juveniles, so that only $(1-d_i)\Delta_+ n_i$ (rather than $\Delta_+ n_i$) territories are secured by type $i$ each iteration. In the heavily disturbed extreme of Grime's triangle, many such bursts (which could each be spatially localized) will occur each iteration (i.e. over the time required for a propagule to reach reproductive maturity), which we approximate as a high constant mortality rate $d\approx 1$. Population persistence requires $b>d/(1-d)$, which implies $b\gg 1$ (severe disturbance comes with a high fecundity/dispersal requirement). We assume this requirement can just be met, so that $b\approx d/(1-d)$ and $L\ll 1$ (Appendix C).

Stressful environments are more ambiguous, and have been the subject of an extensive debate in the plant ecology literature (the ``Grime-Tilman'' debate; \citealt{aerts_1999} and references therein). Severe stress inhibits growth and reproduction, so that $b\ll 1$. In Grime's view, this means that the rate at which propagules successfully develop to adulthood cannot appreciably exceed the mortality rate ($b/d\approx 1$); it is just possible to cope with the environmental stress \citep{grime_1974,grime_1977}. In our model, this implies $L\ll 1$.

The alternative view is that stressed environments are highly competitive. In spite of inhibited growth and reproduction, density is actually high relative to carrying capacity of the stressful environment \citep{taylor_1990}. For example, if consumable resources are scarce, we expect intense resource competition \citep{davis_1998}. Thus, even though $b\ll 1$, we still have $b/d\gg 1$ and $L\ll 1$.

\begin{figure}
\centering
\begin{tabular}{l*{4}{c}}
  & Ideal & Disturbance* & Stress (G) & Stress (HD) \\ \hline
  Constraints & $d \ll 1$ & $d \approx 1$ & $b \ll 1$ & $b \ll 1$ \\
  Other parameters & $b/d\gg 1$ & $(1-d)b\approx d$ & $b\approx d$ & $b/d\gg 1$ \\
  Propagule density $L$  & $\gg 1$ & $\ll 1$ & $\ll 1$ & $\gg 1$
\end{tabular}
\caption{\label{fig:table} The mathematical representation of Grime's three environmental extremes in our model, as well as the high-density (HD) variant of the stressful environment. *Mortality affects both adults and juveniles in the disturbed environment, with $\Delta_+ n_i$ replaced by $(1-d_i)\Delta_+ n_i$.}
\end{figure}

The expected direction of evolution is determined by three factors. First, there may be pleiotropy between $b$, $c$, and $d$, which could be due to physiological trade-offs between them. Second, adaptive mutations in one trait may occur more frequently than in others \citep{yampolsky_2001}. Third, selection favors some trait changes over others: $b$, $c$ and $d$ will tend to evolve in the direction of greatest fitness increase because the probability that an adaptive mutant survives its initial low abundance phase and begins to grow deterministically rather than be lost to drift is proportional to its expected absolute growth rate \citep{haldane_1927,uecker_2011}.

Accounting for the first two factors is beyond the scope of this article. We only consider how selection shapes the direction of evolution for $b$, $c$ and $d$, assuming that mutations affect them independently and with comparable distributions of beneficial fitness effects. Specifically, we consider the effect of mutations that change $b$, $c$, or $d$ by a fixed proportion $b_j=(1+\epsilon) b_i$, $c_j=(1+\epsilon) c_i$ or $d_j=(1-\epsilon) d_i$, where $j$ is is a novel mutant (Appendix C). Mutations with magnitude $\epsilon$ are assumed to occur with the same frequency for all traits.

Not surprisingly, mutations which increase $c$ are effectively neutral when $L\ll 1$ and adaptive when $L\gg 1$. However, selection in our density-dependent lottery model does not favor improvement in $b$ versus $d$. The same is true for proportional $c$-mutations $c_j=(1+\epsilon) c_i$ when $L\gg 1$ in ideal and stressed (HD) environments.

In disturbed environments, where $\Delta_+ n_i$ is replaced by $(1-d_i)\Delta_+ n_i$, the argument for equal selection on $b$ and $d$ is more subtle and contingent. The bulk of the high mortality $d\approx 1$ is environmental and unavoidable. This implies that $\epsilon$ mutations in the remaining genetic part of mortality have a reduced impact on total $d$, suggesting the primacy of $b$ under disturbance. However, this effect is easily matched by the fact that $d_i\approx 1$ and that disturbance mortality affects both juvenile recruitment and adult mortality. In Appendix C, we argue that $b$ and $d$ mutations should contrive to have comparable effects on fitness in the disturbance case (assuming that $d$ can evolve at all) because the fitness benefit of improving $d$ declines as $d$ improves.

%How does Fig.~4 compare to empirical analyses of Grime's C/S/R strategies? In our comparison we will stick to fishes, corals and plants, for which three-way primary strategy schemes are well developed \citep{grime_1977,winemiller_1992,darling_2012}. The connection of our model to fish strategies is necessarily more tentative, given that fishes are motile and not all territorial. 
%
%In disturbed environments, we predict evolution for higher $b$ and lower $d$, but not higher $c$. Higher $b$ means higher fecundity and/or dispersal ability (see ``Model''). This is consistent with a ruderal strategy. Plant ruderals devote a large proportion of their productivity to seed production \citep{grime_1977}, whereas the analogous ``opportunist'' fishes have large intrinsic growth rates \citep{winemiller_1992}. In corals, ruderals are distinguished by brood spawning (rather than broadcast spawning; \citealt{darling_2012}). This corresponds to higher parental investment and lower overall propagule production --- counter-intuitively, a stereotypical ``$K$-selected'', high-density trait \citep{pianka_1970}. However, since broadcast spawners are vulnerable to an Allee effect at the egg fertilization stage \citep{knowlton_2001}, brood spawning could actually be a way to ensure high $b$ at low densities \citep{darling_2012}. Lower $d$ could be achieved by improved individual resistance to physical destruction, but it is hard to reduce mortality in the face of severe disturbances. Alternatively, shortening the time to reproductive maturity (the iteration time in our model) is an effective way of reducing the chance of death per iteration, $d$, for a given frequency of disturbance. An exceptionally short life cycle is probably the most defining characteristic of ruderals \citep{grime_1977,winemiller_1992,darling_2012}. Note that if evolution manages to appreciably reduce $d$ for a given disturbance intensity, then the population no longer lies at the extreme disturbance vertex of Grime's triangle. Thus, ruderals are characterized by both high $b$ and high $d$, but there is a constant pressure for a shorter life cycle, extending the limits of disturbance that can be tolerated. 
%
%In stressful environments, we predict evolution for lower $d$, and also  for higher $c$ in the high-density interpretation of stressful environments. Low $d$ is essential when $b\ll 1$, and stress tolerant plants and corals have long life spans, allowing for long intervals between successful recruitments (and episodic broadcast spawning in corals). For fishes, the ``equilibrium'' strategy is the analogue of Grime's stress tolerator. This strategy is associated with consumable resource limitation, and is also characterized by long life span, as well as high parental investment in tiny broods. This may reflect a high-$c$ strategy in the face of intense competition for severely limited resources (the high-density interpretation).
%
%In ideal environments, we predict evolution for higher $b$ and $c$, but not lower $d$. In plants and corals, a key mechanism for winning territorial contests (higher $c$) is rapidly outgrowing and ``shading out'' competitors; empirically, rapid individual growth is a defining feature of the competitor trait cluster \citep{grime_1977,darling_2012}. The situation for $b$ is more ambiguous. Competitor strategies in plants and corals span a range of $b$. For fishes, the analogous ``periodic'' strategy is characterized by enormous spawn sizes as well as rapid development \citep{winemiller_1992,winemiller_2015}. The role of $b$ in ideal environments will be discussed further below.

@@

%Disentangling the storage and bounded population size effects is not straightforward, and requires a more detailed discussion of each effect than we have space for here. Our model is well suited for such a disentangling since it extends the canonical example of the storage effect to allow for density-dependent effects.

@@

%While adaptive evolution in the direction predicted by our model does produce traits broadly consistent with Grime's scheme, trait data on selection for higher $b$ at high density was ambiguous. This prediction is also counter to the expectations of MacArthur's $r$/$K$ dichotomy \citep{macarthur_1967}, since $b$ is closely related to the maximal, low-density growth rate $r=b-d$ \citep{pianka_1972}, yet in the $r$/$K$ scheme, high density populations should be subject to $K$, not $r$, selection. Yet it is not surprising that $b$ can matter at high densities. In our model (or any lottery model of competition), $b$ matters at high densities because territorial contests among juveniles are intrinsically unpredictable. This is a realistic feature of the model. Even if one genotype is guaranteed to win a territory in a ``fair'' contest (e.g. it is the most efficient exploiter of a limiting consumable resource; \citealt{tilman_1982}), inferior competitors can win by chance. For example, an inferior competitor's propagules may happen to arrive first, gaining a decisive developmental advantage. First arrivals are more likely to occur for genotypes with a fecundity and/or dispersal advantage, as represented by higher $b$ in lottery models. The analogous intuition in the Wright-Fisher model is that fecundity confers a relative fitness advantage, even though population size is not changing. The logistic model for which $r$ and $K$ are named, does not capture this intuition.

@@

Our realization of Grime's triangle (Fig.~\ref{fig:axes}) differs from approaches which identify primary strategies as trait combinations that can stably co-exist \citep{bolker_99}, referring instead to the direction and rate of ongoing adaptive trait evolution under different regimes of stress and disturbance, which is closer in spirit to Grime's arguments \citep{grime_1974,grime_1977}. Moreover, our formulation is mathematical, in contrast to Grime's original verbal and descriptive approach, which is a recognized hindrance to the evaluation or broader application of the C/S/R scheme (e.g. \citealt{tilman_2007}). However, section ``Primary strategies and Grime's triangle'' suggests that, apart from the obvious irrelevance of territorial contests at low density, selection alone has no preference for fecundity, competitive ability or longevity. One or both of the factors we set aside --- pleiotropy/trade-offs and mutation bias --- are needed to get  true ``primary strategy'' trait differentiation.

>>>>

Fitness is typically represented in heavily simplified terms in evolutionary genetics, often using constant selection coefficients

This excludes fundamental ecological factors such as dynamic population size or density-dependence from the most genetically-realistic treatments of evolution, a problem that inspired MacArthur's influential but problematic $r$/$K$ theory.

Following the spirit of $r$/$K$-selection as a general-purpose theory of density-dependent selection

new model of density-dependent selection by generalizing the fixed-density classic lottery model of territorial acquisition to accommodate arbitrary population densities.

We show that, with density dependence, co-existence is possible in the lottery model in a stable environment.

Inspired by natural \textit{Drosophila} populations, we consider co-existence under strong, seasonally-fluctuating selection coupled to large cycles in population density, and show that co-existence (stable polymorphism) is promoted via a combination of the classic storage effect and density-regulated population growth.

@@

@@

In models of genetic evolution, genotypes are typically assigned constant (or occasionally frequency-dependent) selection coefficients describing the change in their relative frequencies over time. This simplified treatment of selection facilitates explicit time-dependent treatment of genotype frequencies, and can be justified over sufficiently short time intervals \citep[p. 276]{ewens_2004}. The emphasis of population genetics is to infer past selection, migration and demographic change given a sample of nucleotide sequences, or to predict how allele frequencies change over time based on their relative fitness effects together with population structure, genetic drift and linkage. The resulting picture of evolution excludes basic elements of the ecological underpinnings of selection, including density dependence, and how selection affects population size. This complicates the inference of past selection, because demographic changes can look genealogically very similar to selective frequency changes \citep{barton_1998}.

@@

 For example, MacArthur's famous $r$/$K$ scheme \citep{macarthur_1962,macarthur_1967} is now almost exclusively known as a framework for understanding life-history traits, and judged on its failure in that role \citep{pianka_1970,stearns_1977,boyce_1984,reznick_2002}. However, the r/K scheme's original purpose was to extend the existing population-genetic treatment of selection to account for population density \citep{macarthur_1962}. Few attempts have been made to develop it further along these lines.

in a simplified, territorial model of growth, dispersal and competition.

We revisit the classic lottery model of \cite{chesson_1981}, which has two features that make it well suited for this role, but one critical flaw that we rectify here.

@@

Taking an example inspired by recent studies of rapid, seasonal evolution in \textit{Drosophila} \citep{bergland_14}, we discuss how environmental fluctuations might stabilize polymorphisms when population density is cyclical.

@@

@@

@@

%This prediction is also counter to the expectations of MacArthur's $r$/$K$ dichotomy \citep{macarthur_1967}, since $b$ is closely related to the maximal, low-density growth rate $r=b-d$ \citep{pianka_1972}, yet in the $r$/$K$ scheme, high density populations should be subject to $K$, not $r$, selection. Yet it is not surprising that $b$ can matter at high densities. In our model (or any lottery model of competition), $b$ matters at high densities because territorial contests among juveniles are intrinsically unpredictable. This is a realistic feature of the model. Even if one genotype is guaranteed to win a territory in a ``fair'' contest (e.g. it is the most efficient exploiter of a limiting consumable resource; \citealt{tilman_1982}), inferior competitors can win by chance. For example, an inferior competitor's propagules may happen to arrive first, gaining a decisive developmental advantage. First arrivals are more likely to occur for genotypes with a fecundity and/or dispersal advantage, as represented by higher $b$ in lottery models. The analogous intuition in the Wright-Fisher model is that fecundity confers a relative fitness advantage, even though population size is not changing. The logistic model for which $r$ and $K$ are named, does not capture this intuition.

@@
